Fixing Vulnerabilities Hinders Maintainability

### 2020-01-31 paper/icpc20.tex
new version

Security is a requirement of utmost importance to produce software with 
high quality. However, there is still a considerable amount of 
vulnerabilities being discovered and fixed, almost weekly. We suspect 
that when developers address these vulnerabilities on their codebases, 
they actually affect the maintainability of software, potentially 
introducing other vulnerabilities. This paper evaluates the impact of 
patches to improve security on the maintainability of open-source software. 
Maintainability is measured based on the Better Code Hub’s model of $10$ 
guidelines using a dataset including $1037$ security commits. \Rui{Add numbers} Results show 
that fixing software vulnerabilities hinders the maintainability of 
open-source software. We conclude that changes to codebases 
to fix vulnerabilities need to be performed with extra care.

@@

implementing refactorings that might affect software maintainability. 

@@

perspective, we leveraged a dataset of $607$ security refactorings collected from open-source 
software. We calculate software maintainability before and after the refactoring to measure its 
impact. This empirical study exhibits proof that changes 

@@

Especially, for refactorings that address \emph{Broken Authentication and Session Management} 

@@

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={JavaStyle}, caption={Fix provided by Netty developers to the
\\CVE-$2014$-$3488$ vulnerability},label={lst:vuln}]

// [snip]
   for (;;) {
+       final int outAppBufSize = engine.getSession().getApplicationBufferSize();
+       final ByteBuffer outAppBuf;
+       if (nioOutAppBuf.capacity() < outAppBufSize) {
+              outAppBuf = ByteBuffer.allocate(outAppBufSize);
+       } else {
+              outAppBuf = nioOutAppBuf;
+       }

@@

In this paper, our concern is to study whether, while improving software

@@

software applications. To answer the following two research questions, we use two datasets

@@

\textit{\textbf{RQ2} What is the impact of security patches versus regular changes on the
\textit{\textbf{RQ3} Which patterns of security patches are more likely to

@@

@@

As said before, the web-based source code analysis service \emph{Better Code
Hub} (BCH) is used to collect the maintainability reports of the refactorings of
each project. Table~\ref{tab:guidelines} presents the $10$ guidelines $-$ proposed
by BCH's authors for delivering software that is not difficult to
maintain~\cite{Visser:2016:OREILLY} $-$ and maps each guideline to the metric 
calculated by BCH. Theses guidelines are calculated using the metrics 
presented in~\cite{criteria:2017}. During each guideline evaluation, the tool
determines the compliance towards one guideline by establishing limits for the
percentage of code allowed to be in each of the $4$ risk severity levels
(\emph{low risk}, \emph{medium risk}, \emph{high risk}, and \emph{very high
risk}). If the project does not violate the thresholds, then it is compliant
with the guideline. \textbf{These thresholds are calibrated by BCH using their own
data/experience --- using open-source and closed software systems}. If a project is
compliant with a guideline, it means that it is at least $65\%$ better than the
software used by BCH to calculate the thresholds\footnote{Check the answer to
\emph{How can I adjust the threshold for passing/not passing a guideline?} at
\url{https://bettercodehub.com/docs/faq} (Accessed on \today{})}.

@@

\begin{figure}[h]	
 	\centering 	\includegraphics[width=0.43\textwidth]{figures/maintainability_sec_vs_reg.pdf}
 	\caption{Maintainability difference for security and baseline patches.}
	\label{fig:secvsreg}

Regarding security commits results, we observe that maintainability decreases in
$41.76\%$ ($433$), remains the same in $19.67\%$ ($204$) and increases in $38.57\%$
($400$) of the cases after security refactorings are applied. The resulting
p-value of the Paired Wilcoxon signed-rank test is $4.78$x$10^{-2}$.
Since the p-value is below the significance level of $0.05$, we argue that
\emph{security pacthes have a negative impact in the maintainability of
open-source software}.

\Sof{Skip, it will be updated soon} For regular commits, we observe that the maintainability decreases in \Sof{<todo>} %$41.35\%$ ($251$) 
of the cases after regular refactorings are applied. But in contrast
to security commits, it increases in \Sof{<todo>} %$58.16\%$ ($353$) 
and remains the same in \Sof{<todo>}
%$0.5\%$ ($3$) 
of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
test is \Sof{<todo>}
%$1.54$x$10^{-6}$
. Since the p-value is below the significance level of $0.05$, 
we argue that \emph{regular refactorings have a
positive impact in the maintainability of open-source software}.

@@

The impact of each type of security patches on the maintainability of
open-source software is presented in Figure~\ref{fig:pat}. The results of the
Paired Wilcoxon signed-rank per pattern exhibited statistical
evidence supporting that the maintainability of open-source software decreases for
only one category: \emph{Miscellaneous} (p-value = $0.033$). Although without 
statistical significance, we also observe that the maintainability of open-source software decreases $3$ different patterns: \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)}, \emph{Information Exposure (CWE-200)} and \emph{Resource Management Errors (CWE-399)}.

@@

@@

\begin{framed}
\textit{\textbf{RQ1} What is the impact of security refactorings on the maintainability
of open-source software?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.1cm}

\textbf{\textit{Security patches harm the maintainability of open-source software.}}
%
We found that $41.76\%$ of the security patches have a negative impact on the
maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). This raises
a new tradeoff when developers need to fix vulnerabilities on their projects
because they might be affecting the software maintainability or even introducing
new vulnerabilities. This study exhibits evidence that developers may have to
reduce maintainability for the sake of security. We understand that developers
should be able to produce secure code without affecting the maintainability of
their projects. But there are still some concerns that need to be addressed:
\begin{itemize}
	\item \textbf{Documentation} of software must provide developers with the
	best practices to implement secure and maintainable software. Unmaintainable
	code is difficult to test, analyze and re-use.

@@

\textbf{\textit{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22), Information Exposure (CWE-200) and Resource Management Errors (CWE-399), and Miscellaneous patterns are more likely to affect
open-source software maintainability.}}\Sof{Posso responder isto sem significancia estatistica?} Although there are a few refactorings
that harm software maintainability in the \emph{Cross-Site Scripting}, overall maintainability does
not suffer any changes for these patterns.

The impact of a refactoring depends on its complexity, i.e., if the refactoring
adds complexity to the code base it is probably affecting the software
maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting}
and \emph{Using Components with Known Vulnerabilities} refactorings endure more
cases without any impact in the maintainability. Typically, these types of
vulnerabilities do not need extra lines to be fixed, as shown in
Listing~\ref{lst:fix}. Whereas, \emph{Broken Authentication} and vulnerabilities
such as the one presented in Section~\ref{sec:motivation} responsible by Denial-of-Service
attacks are more difficult to fix (several code lines deleted and added).
Another example is the Apple authentication flaw discovered in $2017$
(CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
\url{https://support.apple.com/en-us/HT208315} (Accessed on \today{})}, where any user
could log in as root with an empty password. Patrick Wardle examined the cause
of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
\url{https://objective-see.com/blog/blog\_0x24.html} (Accessed on \today{})} and
concluded that the flaw was due to the introduction of high cyclomatic complexity
in a method used to verify the password. In sum, \emph{Cross-Site Scripting} and
\emph{Using Components with Known Vulnerabilities} should take part of the
developers' preoccupations along with the other patterns that explicitly hinder
maintainability because despite their low complexity these are patterns that
still produce a considerable negative impact on the maintainability. Real
efforts have been made in creating and incorporating coding standards and best
practices in software production. However, security is still far from perfection
and fixing vulnerabilities might hinder software maintainability. Thus, it is
not only important to provide tool support to developers to help them apply these
patterns without harming software maintainability, but also to create and redesign
programming languages to facilitate developers improving software maintainability.

\section{Threats to Validity}\label{sec:threats}
%
The following section presents the potential threats to the validity of this
study.
%
\\\textit{Construct:} The maintainability formula was inferred based on the BCH's reports. The high
amount of different projects and backgrounds may require other
maintainability standards. However, BCH does use a representative benchmark of
closed and open-source software projects to compute the
thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
%
\\\textit{Internal:} The security refactorings dataset provided by the previous
work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
commits produced by project developers to classify the changes performed by them
to fix security flaws. This approach discards security refactorings that were
not explicit in commits messages. Baseline commits are retrieved randomly by selecting one commit from the same
project of a given security refactoring. This approach softens the differences
that may result from the characteristics of each project. However,
maintainability may still be affected by the developers' experience, coding
style and software contribution policies which is not evaluated in this study.
Furthermore, this evaluation considers that $607$ regular commits are enough to
alleviate random irregularities in the maintainability differences of the
baseline.
%
\\\textit{External:} We use a dataset that only comprises refactorings of open-source software.
However, the methodology requires to access data that is not publicly available.
Thus, our findings may not extend to non-open source software. Different programming 
languages may require different coding practices to address software safety. The 
dataset comprises more commits in C and PHP, i.e., the dataset may not be representative 
of the population regarding programming languages. Our approach does not consider 
security refactorings in any other languages but English.

@@

This study also shows evidence that there are security patterns that need more
attention than others. \emph{Broken Autentication} issues,
\emph{Denial-of-Service} attacks and \emph{Memory Leaks} endure maintainability
in $55.3\%$, $51.3\%$, and $47.5\%$, respectively. Therefore, we draw attention
for the need of better documentation explaining how to use best practices to
produce secure and maintainable code; new and better programming languages; new
academic curricula in the fields of security and software quality; and, tools
such as BCH to help developers predict the effect of their patches.

As future work, this study may be extended in many directions: analyze which
guidelines are more likely to be affected by security-related refactorings;
investigate what are the most frequent vulnerabilities in each programming
language; investigate how refactorings impact maintainability for each
vulnerability and programming language;
study if there is any correlation between code maintainability and
the popularity of a software project; expand our methodology with other
software quality properties; validate these findings with closed or private
software; and, expand this analysis to other quality standards.

>>>>

Security is a requirement of utmost importance to produce software with 
high quality. However, there is still a considerable amount of 
vulnerabilities being discovered and fixed, almost weekly. We suspect 
that when developers address these vulnerabilities on their codebases, 
they actually affect the maintainability of software, potentially 
introducing other vulnerabilities. This paper evaluates the impact of 
patches to improve security on the maintainability of open-source software. 
Maintainability is measured based on the Better Code Hub’s model of $10$ 
guidelines using a dataset including $1037$ security commits. Results show 
that fixing software vulnerabilities hinders the maintainability of 
open-source software. This is particularly the case for patches that address Path Traversal (CWE-22),
Information Exposure (CWE-200) and Resource Management Errors (CWE-399) . We conclude that changes to code bases 
to fix vulnerabilities need to be performed with extra care.

@@

implementing patches that might affect software maintainability. 

@@

perspective, we leveraged a dataset of $1335$ security patches collected from open-source 
software. We calculate software maintainability before and after the patch to measure its 
impact. This empirical study presents evidence that changes 

@@

Especially, for patches that address \emph{Broken Authentication and Session Management} 

@@

\medskip
\setcounter{lstannotation}{0}
\begin{lstlisting}[style={JavaStyle}, caption={Fix provided by Netty developers to the
\\CVE-$2014$-$3488$ vulnerability},label={lst:vuln}]
// [snip]
   for (;;) {
+       final int outAppBufSize = engine.getSession().getApplicationBufferSize();
+       final ByteBuffer outAppBuf;
+       if (nioOutAppBuf.capacity() < outAppBufSize) {
+              outAppBuf = ByteBuffer.allocate(outAppBufSize);
+       } else {
+              outAppBuf = nioOutAppBuf;
+       }

@@

Software maintainability is designed as the degree to which an application is understood, repaired or enhanced. In this paper, our concern is to study whether, while improving software

@@

software applications. This is important because software maintainability is approximately $75\%$ of the cost related to a project. To answer the following three research questions, we use two datasets

@@

\textit{\textbf{RQ2} Which patterns of security patches are more likely to

@@

\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the

@@

In this research, we follow a very similar methodology to the one presented in previous work on the maintainability 
of energy-oriented fixes~\cite{cruz2019energyoriented}. As said before, the web-based source code analysis service \emph{Better Code
Hub} (BCH) is used to collect the maintainability reports of the refactorings of
each project. Table~\ref{tab:guidelines} presents the $10$ guidelines $-$ proposed
by BCH's authors for delivering software that is not difficult to
maintain~\cite{Visser:2016:OREILLY} $-$ and maps each guideline to the metric 
calculated by BCH. Theses guidelines are calculated using the metrics 
presented in~\cite{criteria:2017}. During each guideline evaluation, the tool
determines the compliance towards one guideline by establishing limits for the
percentage of code allowed to be in each of the $4$ risk severity levels
(\emph{low risk}, \emph{medium risk}, \emph{high risk}, and \emph{very high
risk}). If the project does not violate the thresholds, then it is compliant
with the guideline. \textbf{These thresholds are calibrated by BCH using their own
data/experience --- using open-source and closed software systems}. If a project is
compliant with a guideline, it means that it is at least $65\%$ better than the
software used by BCH to calculate the thresholds\footnote{Check the answer to
\emph{How can I adjust the threshold for passing/not passing a guideline?} at
\url{https://bettercodehub.com/docs/faq} (Accessed on \today{})}.

@@

@@

The impact of each type of security patches on the maintainability of
open-source software is presented in Figure~\ref{fig:pat}. The results of the
Paired Wilcoxon signed-rank per pattern exhibited statistical
evidence supporting that the maintainability of open-source software decreases for
only one category: \emph{Miscellaneous} (p-value = $0.033$). Although without 
statistical significance, we also observe that the maintainability of open-source software decreases in $3$ different patterns: \emph{Improper Limitation of a Pathname to a Restricted Directory/Path Traversal (CWE-22)}, \emph{Information Exposure (CWE-200)} and \emph{Resource Management Errors (CWE-399)}.

@@

\begin{framed}
\textit{\textbf{RQ3} What is the impact of security patches versus regular changes on the
maintainability of open-source software?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.1cm}

\begin{figure}[h]	
 	\centering 	\includegraphics[width=0.43\textwidth]{figures/maintainability_sec_vs_reg.pdf}
 	\caption{Maintainability difference for security and baseline patches.}
	\label{fig:secvsreg}
\end{figure}

Regarding security commits results, we observe that maintainability decreases in
$41.76\%$ ($433$), remains the same in $19.67\%$ ($204$) and increases in $38.57\%$
($400$) of the cases after security patches are applied. The resulting
p-value of the Paired Wilcoxon signed-rank test is $4.78$x$10^{-2}$.
Since the p-value is below the significance level of $0.05$, we argue that
\emph{security patches have a negative impact in the maintainability of
open-source software}.

\Sof{Skip, it will be updated soon} For regular commits, we observe that the maintainability decreases in \Sof{<todo>} %$41.35\%$ ($251$) 
of the cases after regular refactorings are applied. But in contrast
to security commits, it increases in \Sof{<todo>} %$58.16\%$ ($353$) 
and remains the same in \Sof{<todo>}
%$0.5\%$ ($3$) 
of the cases. The resulting p-value of the Paired Wilcoxon signed-rank
test is \Sof{<todo>}
%$1.54$x$10^{-6}$
. The baseline is still being computed by BCH but we can already see
that regular changes usually have no impact on the maintainability of
software.

%Since the p-value is below the significance level of $0.05$, 
%we argue that \emph{regular refactorings have a
%positive impact in the maintainability of open-source software}.

@@

\textbf{\textit{Security patches harm the maintainability of open-source software.}}
%

\textbf{\textit{Path Traversal (CWE-22), Information Exposure (CWE-200), Resource Management Errors (CWE-399) are more likely to affect
open-source software maintainability.}}

Overall, the results obtained for the maintainability per CWE are no statistical significance

The impact of a patch depends on its complexity, i.e., if the patchs
adds complexity to the code base it is probably affecting the software
maintainability. One evidence of that is the fact of \emph{Cross-Site Scripting} (CWE-79)
patches endure more
cases without any impact in the maintainability. Typically, these types of
vulnerabilities do not need extra lines to be fixed, as shown in
Listing~\ref{lst:fix}. Whereas, \emph{Resource Management Errors} (CWE-399) and vulnerabilities
such as the one presented in Section~\ref{sec:motivation} responsible by Denial-of-Service
attacks are more difficult to fix (several code lines deleted and added).
Another example is the Apple authentication flaw discovered in $2017$
(CVE-$2017$-$13872$)\footnote{CVE-$2017$-$13872$ details available at
\url{https://support.apple.com/en-us/HT208315} (Accessed on \today{})}, where any user
could log in as root with an empty password. Patrick Wardle examined the cause
of the issue\footnote{\emph{Why $<$blank$>$ Gets You Root} available at
\url{https://objective-see.com/blog/blog\_0x24.html} (Accessed on \today{})} and
concluded that the flaw was due to the introduction of high cyclomatic complexity
in a method used to verify the password. In sum, \emph{Cross-Site Scripting} (CWE-79) should take part of the
developers' preoccupations along with the other patterns that explicitly hinder
maintainability because despite their low complexity these are patterns that
still produce a considerable negative impact on the maintainability. Real
efforts have been made in creating and incorporating coding standards and best
practices in software production. However, security is still far from perfection
and fixing vulnerabilities might hinder software maintainability. Thus, it is
not only important to provide tool support to developers to help them apply these
patterns without harming software maintainability, but also to create and redesign
programming languages to facilitate developers improving software maintainability.

\begin{framed}
\textit{\textbf{RQ3} What is the impact of security patches versus regular changes
on the maintainability of open-source software?}
\vspace{-0.1cm}
\end{framed}
\vspace{-0.1cm}

We found that $41.76\%$ of the security patches have a negative impact on the
maintainability of open-source software (cf. Figure~\ref{fig:secvsreg}). This raises
a new trade-off when developers need to fix vulnerabilities on their projects
because they might be affecting the software maintainability or even introducing
new vulnerabilities. This study exhibits evidence that developers may have to
reduce maintainability for the sake of security. We understand that developers
should be able to produce secure code without affecting the maintainability of
their projects. But there are still some concerns that need to be addressed:
\begin{itemize}
	\item \textbf{Documentation} of software must provide developers with the
	best practices to implement secure and maintainable software. Unmaintainable
	code is difficult to test, analyze and re-use.

@@

\section{Threats to Validity}\label{sec:threats}
%
The following section presents the potential threats to the validity of this
study.
%
\\\textit{\textbf{Construct:}} The maintainability formula was inferred based on the BCH's reports. The high
amount of different projects and backgrounds may require other
maintainability standards. However, BCH does use a representative benchmark of
closed and open-source software projects to compute the
thresholds for each maintainability guideline~\cite{Visser:2016:OREILLY, Baggen2012}.
%
\\\textit{\textbf{Internal:}} The security patches dataset provided by the previous
work~\cite{Reis:2017:IJSSE} was collected based on the messages of GitHub
commits produced by project developers to classify the changes performed by them
to fix security flaws. This approach discards security patches that were
not explicit in commits messages. Baseline commits are retrieved randomly by selecting one commit from the same
project of a given security patch. This approach softens the differences
that may result from the characteristics of each project. However,
maintainability may still be affected by the developers' experience, coding
style and software contribution policies which is not evaluated in this study.
Furthermore, this evaluation considers that $1037$ regular commits are enough to
alleviate random irregularities in the maintainability differences of the
baseline.
%
\\\textit{\textbf{External:}} We use a dataset that only comprises patches of open-source software.
However, the methodology requires to access data that is not publicly available.
Thus, our findings may not extend to non-open source software. Different programming 
languages may require different coding practices to address software safety. The 
dataset comprises more commits in Java, i.e., the dataset may not be representative 
of the population regarding programming languages. Our approach does not consider 
security patches in any other languages but English.

@@

This study also shows evidence that there are security patterns that need more
attention than others. \emph{Path Traversal} (CWE-22),
\emph{Information exposure} (CWE-200) and \emph{Resource Management Errors} (CWE-399) vulnerabilities endure maintainability
in $48.6\%$, $51.2\%$, and $54.5\%$, respectively. Therefore, we draw attention
for the need of better documentation explaining how to use best practices to
produce secure and maintainable code; new and better programming languages; new
academic curricula in the fields of security and software quality; and, tools
such as BCH to help developers predict the effect of their patches.

As future work, this study may be extended in many directions: 
investigate what are the most frequent vulnerabilities in each programming
language; investigate how patches impact maintainability for each
vulnerability and programming language;
study if there is any correlation between code maintainability and
the popularity of a software project; expand our methodology with other
software quality properties; validate these findings with closed or private
software; and, expand this analysis to other quality standards.
